Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2018-12-11 scagnostics.Rmd
pipe section plus spell check

[@Grimm2016] explored the behaviour of scagnostics for selecting variables, and proposed two more, based on smoothing splines and distance correlation, that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518] round out our collection of metrics. \footnote{It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in \cite{huber85,jones87}.}
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

The paper investigates the behaviour of these newly-defined indexes. Section \ref{sec:construct} discusses index construction, and how they fit into the guided tour. Section \ref{sec:investigate} investigates the behaviour of the indexes, particularly in relation to optimisation. The new guided tour methods are applied to an example where the posterior distributions from physics models (Section \ref{sec:phys}) are explored.

@@

- Definition of pp index
- Optimisation
- Illustration of what is already available in tourr
- New index explanations
-->
A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data distribution. Typically the definition is such that larger values of $f$ indicate more interesting distribution, and therefore maximising $f$ over all possible $d$ dimensional projections of a data set with $p>d$ parameters will find the most interesting projections. As most indices characterise deviations from a normal distribution, one would first map the empiric data distribution onto a density for which such deviations can then be defined in different ways, see e.g. @CBC93. When departing from the idea of characterising differences to the normal distribution it may however be preferred to work directly with the empirical data distribution and instead characterise it by measures based on e.g. the minimal spanning tree or the mutual information.

## Standardization
Lower dimensional projections are highly sensitive to standardization performed on the original distribution. We want to avoid e.g.  highlighting directions based only on different scales, or artifically extending the parameter range by including outliers, which may result in seemingly linear behaviour for most points. Methods to consider include rescaling of individual directions to a common interval (e.g. to fall between $[0,1]$), sphering, outlier removal or transformation to a logarithmic scale. The method(s) of choice should be informed by the distributions found in the data set as well as aims of the analysis. \textcolor{red}{particular challenges: noise only directions, identification of outliers?}
\textcolor{red}{for holes/cmass index they are only well defined when data is standardised, how about the other index functions?}
\textcolor{red}{When keeping in mind the concrete example of posterior samples outliers should not be an issue, if extreme outliers are found one should probably investigate origin. Also noise only directions, e.g. nuissance parameters, are also not an issue as they contain no structure, see BBH example.}

\textcolor{red}{\cite{jones87} discusses centering and sphering of the data points. Centering is needed if index is not translational invariant (should not be the case for any index considered here). Arguments for sphering are that scale effects should be extracted, we don't want to reproduce information already obtained by PCA. To my mind rescaling to an overall intervall is more appropriate here. Rotations between parameters should not affect the outcome, but will make interpretation more difficult. Standardising the variance will remove information about the posterior variance. Overall we are interested in strong association only, if these structures were affected by the standardisation procedure, it would most likely be unwelcome effects?}

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuatins or a consequence of noisy index functions, and @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimisation for the identification of the maximum. On the other hand the global maximum may not be the only projection of interest, and one may want to observe the selected views in the context of the full distribution rather than a static view. These issues are addressed when combining projection pursuit with the grand tour [@CBCH94]. In this case the properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshoothing and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible alogrithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. In addition the tourr package provides the \texttt{search\_geodesic} search function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimises the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

\textcolor{red}{\cite{posse95b} discusses the optimisation, in particular that for most index functions and optimisers results are too local, largely dependent on starting point. Fiedmans suggestion of hybrid steppind search + steepest-ascent still remains fairly local. \cite{posse90} suggested random search instead, which seems to perform better, but did not find pdf of that article, so not so sure what he actually does.}

## New index functions
\label{sec:indexDef}
Focusing now on $d=2$ dimensional projections, we aim to identify projections that reveal interesting structures or correlations between combinations of variables. We therefore draw on index functions that have been developped for the selection of variable pairs that show interesting features in a scatter plot, but we generalise the notion such that instead of variable pairs we consider any $2$ dimensional projection of the multivariate data distribution. It means that we consider the projection axes as input to the index functions, which will thus score the interestingness of the given projection. What we are optimising over are then the entries of the $p\times2$ dimensional projection matrix, considering that they obey orthonormality conditions.
Concretely here we consider the following types of index functions:\footnote{We note that all descriptions are idealized, while in practice approximations and binning are used to obtain reasonable computing time, we refer to original references and package documentation for details.}

* From the scagnostics family select those best suited to detect shapes, with definitions given below. They are defined to take values between [0,1]. \textcolor{red}{The definition of these should be rotation invariant, however, some small dependence is introduced via the hex binning that is done before calculating the scagnostics indexes.}
    + Convex: The convex measure is defined as $c_{convex}= \frac{area(A)}{area(H)}$ where $A$ is the alpha hull and $H$ the convex hull. This is the only measure where interesting projections will take low values and one may either minimize the index function or maximize $1-c_{convex}$.
    + Skinny: The skinnyness of a polygon can be measured as the ratio of the perimeter to the area. We define $c_{skinny} = 1 -  \frac{\sqrt{4\pi area(A)}}{perimeter(A)}$, where the normalisation is chosen such that $c_{skinny} = 0$ for a full circle, and values close to one or skinny polygons.
    + Stringy: The stringy index is a measure of the branching structure of the minimal spanning tree (MST), $c_{stringy} = \frac{diameter(MST)}{length(MST)}$ where the diameter is the longest connected path, and the length is the total length (sum of all edges), i.e. $c_{stringy}=1$ if the MST contains no branches.

* New index functions available in @mbgraphic, defined to fall between [0,1] \textcolor{red}{and not rotation invariant, as I think Var(X)*Var(Y) introduces rotation invariance even for distance correlation?}:
    + Distance correlation defined in @szekely2007: The definition is based on empiric measures of distance covariance and variance defined on the dataset 
    \begin{equation}
(X,Y) = \{(x_1,y_1), ..., (x_n, y_n) \in \mathbb{R}^p \times \mathbb{R}^q\}
\end{equation}
 as
 \begin{equation}
 Cov^2(X,Y) = \frac{1}{n^2} \sum_{k,l=1}^{n} A_{kl}B_{kl},\\
 Var^2(X) = \frac{1}{n^2} \sum_{k,l=1}^{n} A_{kl}^2
 \end{equation}
 where $A_{kl}=a_{kl}-\bar{a}_{k.}-\bar{a}_{l.}+\bar{a}_{..}$ with
 \begin{equation}
 a_{kl} = | x_k - x_l|_p,\\
 \bar{a}_{k.} = \frac{1}{n} \sum_{l=1}^n a_{kl},\\
  \bar{a}_{.l} = \frac{1}{n} \sum_{k=1}^n a_{kl},\\
 \bar{a}_{..} = \frac{1}{n^2}\sum_{k,l=1}^n a_{kl}
 \end{equation}
 and reads
 \begin{equation}
 dCor(X,Y) = \begin{cases}
 \frac{Cov(X,Y)}{\sqrt{Var(X)Var(Y)}} & Var(X)Var(Y) > 0\\
 0 & Var(X)Var(Y) = 0
 \end{cases}
 \end{equation}
    + Spline based measure: Given a spline model of the data distribution that is obtained via the R implentation in the \texttt{gam} function in the mgcv R package [@w16], we define the index based on the variance of the residuals:
    \begin{equation}
    splines_{2d} = max(1- \frac{Var(res_{X\sim Y})}{Var(X)}, 1-\frac{Var(res_{Y\sim X})}{Var(Y)}),
    \end{equation}
    which takes large values if the distribution is well described by the spline model, indicating functional dependence.

* Information based index functions from @Reshef1518 are based on the mutual information $I$ when considering the distribution induced by the data points $D \subset \mathbb{R}^2$ on an $x-$by$y$ grid $G$. With $I^{\ast} (D, x, y) = \max I (D|_G)$ the characteristic matrix $M$ is defined as
\begin{equation}
M(D)_{x,y} = \frac{I^{\ast} (D, x, y)}{\log \min{x,y}},
\end{equation}
where the denominator is chosen such that each entry of $M$ will take values between [0,1].
We can now define the following metrics:
    + Maximum Information Coefficient (MIC):
    \begin{equation}
    MIC(D) = \max_{xy<B(n)}\{M(D)_{x,y}\},
    \label{eq:MIC}
    \end{equation}
    where by default $B(n) = n^{0.6}$
    + Total Information Coefficient (TIC), see @JMLRv1715308:
    \begin{equation}
    TIC(D) = \sum_{xy<B(n)} M(D)_{x,y}
    \end{equation}
  While $MIC$ is normalised to fall between $[0,1]$, the upper bound of $TIC$ depends on the sample size via the bounding function $B(n)$. $TIC$ was designed to be more stable test against null hypothesis of independence, using the full information of the characteristic matrix rather than just the maximum, and as such is expected to be better suited as a ppi.

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers (\cite{diaconis84}, \cite{huber85}, \cite{jones87}, \cite{posse95b}, \cite{hall89}). The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modeling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimisation feasible, small angles between projections correspond to small perturbations of the sample, and thus shold be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the crteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.

@@

- **squintability**: \cite{barnett1981interpreting} introduced the idea of squint angle to indicate resolution of structure in the data. Fine structure like the parallel planes in the infamous RANDU data \cite{Marsaglia68} has a small squint angle because you have to be very close to the optimal projection plane to be able to see the structure. Structures with small squint angle are difficult to find, because the optmisation algorithm needs to get very close to begin hill-climbing to the optimum. The analyst doesn't have control over the data structure, but does have control over the PPI. Squintability is about the shape of the PPI over all projections. It should have smooth low values for noise projections and a clearly larger value, ideally with a big squint angle, for structured projections. The optimiser should be able to clearly see the optimal projections as different from noise. To examine squintability, the PPI values are examined on interpolated tour paths between a noise projection and a distant structured projection. 

@@

- **rotation invariance**: The orientation of structure within a projection plane should not change the index value. This is especially important when using the projection pursuit guided tour, because the tour path is defined between planes, along a geodesic path, not bases within planes. If a particular orientation is more optimal, this will get lost as the projection shown pays no attention to orientation. \cite{Buja2004ComputationalMF} describes alternative interpolation paths based on Givens and Householder rotations which progress from basis to basis. It may be possible to ignore rotation invariance with these interpolations but there isn't a current implementation, primarily because the within-plane spin that is generated is distracting from a visualisation perspective. Rotation invariance is checked for the proposed PPIs by rotating the structured projection, within the plane. <!--

@@

- **speed**: Being fast to compute allows the index to be used in real-time in a guided tour, where the optimisation can be watched. When the computations are shifted off-line, to watch in replay, computation times matter less. This is checked  by comparing times for benchmark scenarios with varying sample size and dimension.

@@

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the PPIs on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $p$ is used for the number of points, $n$ is the number of dimensions, and $d=2$ is the projection dimension.

@@

The structured projection is in variables 5 and 6 ($x_5, x_6$). Two samples sizes are used: $n=(100, 1000)$. All variables are standardized to have mean 0 and standard deviation 1. Figure \ref{fig:data} shows samples from each of the families, of the nuisance and structured pairs of variables. Table \ref{tab:indexTable} compares the PPIs for structured projections against those for nuisance variables, based on 100 simulated data sets of each type, using sample size 1000. The lower and upper show the 5th and 95th percentile of values. All indexes, except convex show distinctly higher values for the structured projections. The convex index shows the inverse scale to other indices, thus (1-convex) will be used in the assessment of performance of PPIs. The scale for the holes index is smaller than the others ranging from about 0.7 through 1, so it is re-scaled in the performace assessment so that all indices can be plotted on a common scale of 0-1 (details in Appendix). Similarly, the TIC index is re-scaled.

@@

<!-- - Compare the index values for structured vs nuisance variables, by simulating many examples from the model families, and recording the 90% confidence intervals. This checks if the PPI can detect the structure, that it has distinct differences between the sets of values. 
-->
1. Compute the PPI values on the tour path along an interpolation between pairs of nuisance variables, $x_1 - x_2$ view either to the $x_3 - x_4$. The result is ideally a smooth change in low values. This checks the smoothness property. 
2. Change to a tour path between a pair of nuisance variables $x_1 - x_2$ and the structured pair of variables $x_5 - x_6$, and compute the PPI along this. This examines the squintability, and smoothness. If the function is smooth and slowly increases towards the structured projection, then the structure is visible from a distance. 
3. Use the guided tour to examine the ease of optimisation. This depends on having a relatovely smooth function, with structure visible from a distance. One index is optimised to show how effectively the maximum is attained, and the values for other PPIs is examined along the same path, to examine the similarity between PPIs. 
4. Rotation invariance is checked by computing PPIs on rotations of the structured projection. 
5. Some PPIs have a choice of parameters, and the choice can have an effect on function smoothness, and sensitivity to structure. The parameters for these PPIs is varied and examined along a tour path between noise and structured projections, to examine the effects of parameter choices and how fiddly the index is to tune. 
6. Computational speed for the selected indices is examined on a range of sample sizes.

@@

Figure \ref{fig:plannedtour} shows the PPIs for a tour sequence between a nuisance and structured projection. Sample size is indicated by linetype: dashed being $n=100$ and solid is $n=1000$. The beginnnig of the sequence is the nuisance projection and the end is the structured projection. The index values for most PPIs increases substantially nearing the structured projection, indicating that they "see" the structure. Some indexes see all three structures: scagnostics, MIC and TIC,  which means that they are flexible indexes capable of detecting a range of structure. [@Grimm2016]'s indexes, dcor2d and splines2d, are excellent for detecting the sine, and they can see it from far away, indicated by the long slow increase in index value. The holes index easily detects the pipe, and can see it from a distance. The scagnostic index, stringy, can see the structure but is myopic, only when it is very close. Interestingly the scagnostic, skinny, sees the spiral from a distance.

@@

The guided tour combines optimisation with interpolation between pairs of planes. Target planes of the path are chosen to maximise the PPI. There are three derivative-free optimisation methods available in the guided tour: \verb# search_better_random# (1), \verb# search_better# (2), and \verb# search_geodesic# (3). Method 1 casts a wide net randomly generating projection planes, computing the PPIs and keeping the best projection, and method 2 conducts a localised maximum search. Method 3 is quite different: a local search is conducted to determine a promising direction, and then this direction is followed until the maximum in that direction is found. For all methods the optimisation is iterative, the best projections form target planes in the tour, the tour path is the interpolation to this target, and then a new search for a better projection is made, followed by the interpolation. For each projection during the interpolation steps, the PPI is recorded.

The stopping rule is that no better projections are found after a fixed number of tries, given a fixed tolerance value measuring difference. For method 1 and 2 two additional parameters control the optimisation: the search window $\alpha$, giving the maximum distance from the current plane in which projections are sampled, and the cooling factor, giving the incremental decrease in search window size. Method 3 in principle also has two free parameters, which are however fixed in the current implementation. The first is the small step size used when evaluating the most promising direction, it is fixed to 0.01, and the second parameter being the window over which the line search is performed, fixed to $\pm \pi/4$ away from the current plane.

For distributions and indexes with smooth behaviour and good squintability, method 3 is the most effective method for optimisation. If these two criteria are not met the method may still be useful, but only given an informed starting projection. In such cases we can follow a method similar to that proposed by Friedman (XXX REF): we break the optimisation in two distinct steps. A first step ("scouting") uses method 2 with large search window and no cooling as a way of stepping over fluctuations and local maxima and yielding an approximation of the global maximum. Note that this likely requires large number of tries, especially as dimension increases, since most randomly picked planes will not be interesting. The second step is uses method 3 starting from the approximate maximum, which will take small steps to refine the result to be closer to the global maximum.

@@

As seen in Figure \ref{fig:plannedtour} the pipe structure has a small squint angle, and indeed using only method 3 does not uncover it in the six dimensional dataset. We therefore adapt the two step procedure described above. For the scouting we use $\alpha = 0.5$ and stopping condition of maximum 5000 tries, and we optimise the TIC index.

>>>>

[@Grimm2016] explored the behavior of scagnostics for selecting variables, and proposed two more, based on smoothing splines and distance correlation, that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518] round out our collection of metrics. \footnote{It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in \cite{huber85,jones87}.}
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

The paper investigates the behavior of these newly-defined indexes. Section \ref{sec:construct} discusses index construction, and how they fit into the guided tour. Section \ref{sec:investigate} investigates the behavior of the indexes, particularly in relation to optimization. The new guided tour methods are applied to an example where the posterior distributions from physics models (Section \ref{sec:phys}) are explored.

@@

- Definition of pp index
- Optimisation
- Illustration of what is already available in tourr
- New index explanations
-->
A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data distribution. Typically the definition is such that larger values of $f$ indicate more interesting distribution, and therefore maximizing $f$ over all possible $d$ dimensional projections of a data set with $p>d$ parameters will find the most interesting projections. As most indices characterize deviations from a normal distribution, one would first map the empiric data distribution onto a density for which such deviations can then be defined in different ways, see e.g. @CBC93. When departing from the idea of characterizing differences to the normal distribution it may however be preferred to work directly with the empirical data distribution and instead characterize it by measures based on e.g. the minimal spanning tree or the mutual information.

## Standardization
Lower dimensional projections are highly sensitive to standardization performed on the original distribution. We want to avoid e.g.  highlighting directions based only on different scales, or artificially extending the parameter range by including outliers, which may result in seemingly linear behavior for most points. Methods to consider include rescaling of individual directions to a common interval (e.g. to fall between $[0,1]$), sphering, outlier removal or transformation to a logarithmic scale. The method(s) of choice should be informed by the distributions found in the data set as well as aims of the analysis. \textcolor{red}{particular challenges: noise only directions, identification of outliers?}
\textcolor{red}{for holes/cmass index they are only well defined when data is standardized, how about the other index functions?}
\textcolor{red}{When keeping in mind the concrete example of posterior samples outliers should not be an issue, if extreme outliers are found one should probably investigate origin. Also noise only directions, e.g. nuisance parameters, are also not an issue as they contain no structure, see BBH example.}

\textcolor{red}{\cite{jones87} discusses centering and sphering of the data points. Centering is needed if index is not translational invariant (should not be the case for any index considered here). Arguments for sphering are that scale effects should be extracted, we don't want to reproduce information already obtained by PCA. To my mind rescaling to an overall interval is more appropriate here. Rotations between parameters should not affect the outcome, but will make interpretation more difficult. Standardizing the variance will remove information about the posterior variance. Overall we are interested in strong association only, if these structures were affected by the standardization procedure, it would most likely be unwelcome effects?}

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuations or a consequence of noisy index functions, and @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimization for the identification of the maximum. On the other hand the global maximum may not be the only projection of interest, and one may want to observe the selected views in the context of the full distribution rather than a static view. These issues are addressed when combining projection pursuit with the grand tour [@CBCH94]. In this case the properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshooting and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible algorithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. In addition the tourr package provides the \texttt{search\_geodesic} search function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimists the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

\textcolor{red}{\cite{posse95b} discusses the optimization, in particular that for most index functions and optimists results are too local, largely dependent on starting point. Fiedmans suggestion of hybrid stepping search + steepest-ascent still remains fairly local. \cite{posse90} suggested random search instead, which seems to perform better, but did not find pdf of that article, so not so sure what he actually does.}

## New index functions
\label{sec:indexDef}
Focusing now on $d=2$ dimensional projections, we aim to identify projections that reveal interesting structures or correlations between combinations of variables. We therefore draw on index functions that have been developed for the selection of variable pairs that show interesting features in a scatter plot, but we generalist the notion such that instead of variable pairs we consider any $2$ dimensional projection of the multivariate data distribution. It means that we consider the projection axes as input to the index functions, which will thus score the interestingness of the given projection. What we are optimizing over are then the entries of the $p\times2$ dimensional projection matrix, considering that they obey orthonormality conditions.
Concretely here we consider the following types of index functions:\footnote{We note that all descriptions are idealized, while in practice approximations and binning are used to obtain reasonable computing time, we refer to original references and package documentation for details.}

* From the scagnostics family select those best suited to detect shapes, with definitions given below. They are defined to take values between [0,1]. \textcolor{red}{The definition of these should be rotation invariant, however, some small dependence is introduced via the hex binning that is done before calculating the scagnostics indexes.}
    + Convex: The convex measure is defined as $c_{convex}= \frac{area(A)}{area(H)}$ where $A$ is the alpha hull and $H$ the convex hull. This is the only measure where interesting projections will take low values and one may either minimize the index function or maximize $1-c_{convex}$.
    + Skinny: The skinnyness of a polygon can be measured as the ratio of the perimeter to the area. We define $c_{skinny} = 1 -  \frac{\sqrt{4\pi area(A)}}{perimeter(A)}$, where the normalization is chosen such that $c_{skinny} = 0$ for a full circle, and values close to one or skinny polygons.
    + Stringy: The stringy index is a measure of the branching structure of the minimal spanning tree (MST), $c_{stringy} = \frac{diameter(MST)}{length(MST)}$ where the diameter is the longest connected path, and the length is the total length (sum of all edges), i.e. $c_{stringy}=1$ if the MST contains no branches.

* New index functions available in @mbgraphic, defined to fall between [0,1] \textcolor{red}{and not rotation invariant, as I think Var(X)*Var(Y) introduces rotation invariance even for distance correlation?}:
    + Distance correlation defined in @szekely2007: The definition is based on empiric measures of distance covariance and variance defined on the dataset 
    \begin{equation}
(X,Y) = \{(x_1,y_1), ..., (x_n, y_n) \in \mathbb{R}^p \times \mathbb{R}^q\}
\end{equation}
 as
 \begin{equation}
 Cov^2(X,Y) = \frac{1}{n^2} \sum_{k,l=1}^{n} A_{kl}B_{kl},\\
 Var^2(X) = \frac{1}{n^2} \sum_{k,l=1}^{n} A_{kl}^2
 \end{equation}
 where $A_{kl}=a_{kl}-\bar{a}_{k.}-\bar{a}_{l.}+\bar{a}_{..}$ with
 \begin{equation}
 a_{kl} = | x_k - x_l|_p,\\
 \bar{a}_{k.} = \frac{1}{n} \sum_{l=1}^n a_{kl},\\
  \bar{a}_{.l} = \frac{1}{n} \sum_{k=1}^n a_{kl},\\
 \bar{a}_{..} = \frac{1}{n^2}\sum_{k,l=1}^n a_{kl}
 \end{equation}
 and reads
 \begin{equation}
 dCor(X,Y) = \begin{cases}
 \frac{Cov(X,Y)}{\sqrt{Var(X)Var(Y)}} & Var(X)Var(Y) > 0\\
 0 & Var(X)Var(Y) = 0
 \end{cases}
 \end{equation}
    + Spline based measure: Given a spline model of the data distribution that is obtained via the R implantation in the \texttt{gam} function in the mgcv R package [@w16], we define the index based on the variance of the residuals:
    \begin{equation}
    splines_{2d} = max(1- \frac{Var(res_{X\sim Y})}{Var(X)}, 1-\frac{Var(res_{Y\sim X})}{Var(Y)}),
    \end{equation}
    which takes large values if the distribution is well described by the spline model, indicating functional dependence.

* Information based index functions from @Reshef1518 are based on the mutual information $I$ when considering the distribution induced by the data points $D \subset \mathbb{R}^2$ on an $x-$by$y$ grid $G$. With $I^{\ast} (D, x, y) = \max I (D|_G)$ the characteristic matrix $M$ is defined as
\begin{equation}
M(D)_{x,y} = \frac{I^{\ast} (D, x, y)}{\log \min{x,y}},
\end{equation}
where the denominator is chosen such that each entry of $M$ will take values between [0,1].
We can now define the following metrics:
    + Maximum Information Coefficient (MIC):
    \begin{equation}
    MIC(D) = \max_{xy<B(n)}\{M(D)_{x,y}\},
    \label{eq:MIC}
    \end{equation}
    where by default $B(n) = n^{0.6}$
    + Total Information Coefficient (TIC), see @JMLRv1715308:
    \begin{equation}
    TIC(D) = \sum_{xy<B(n)} M(D)_{x,y}
    \end{equation}
  While $MIC$ is normalized to fall between $[0,1]$, the upper bound of $TIC$ depends on the sample size via the bounding function $B(n)$. $TIC$ was designed to be more stable test against null hypothesis of independence, using the full information of the characteristic matrix rather than just the maximum, and as such is expected to be better suited as a PPI.

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers (\cite{diaconis84}, \cite{huber85}, \cite{jones87}, \cite{posse95b}, \cite{hall89}). The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modeling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimization feasible, small angles between projections correspond to small perturbations of the sample, and thus should be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the criteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.

@@

- **squintability**: \cite{barnett1981interpreting} introduced the idea of squint angle to indicate resolution of structure in the data. Fine structure like the parallel planes in the infamous RANDU data \cite{Marsaglia68} has a small squint angle because you have to be very close to the optimal projection plane to be able to see the structure. Structures with small squint angle are difficult to find, because the optimization algorithm needs to get very close to begin hill-climbing to the optimum. The analyst doesn't have control over the data structure, but does have control over the PPI. Squintability is about the shape of the PPI over all projections. It should have smooth low values for noise projections and a clearly larger value, ideally with a big squint angle, for structured projections. The optimizer should be able to clearly see the optimal projections as different from noise. To examine squintability, the PPI values are examined on interpolated tour paths between a noise projection and a distant structured projection. 

@@

- **rotation invariance**: The orientation of structure within a projection plane should not change the index value. This is especially important when using the projection pursuit guided tour, because the tour path is defined between planes, along a geodesic path, not bases within planes. If a particular orientation is more optimal, this will get lost as the projection shown pays no attention to orientation. \cite{Buja2004ComputationalMF} describes alternative interpolation paths based on Givens and Householder rotations which progress from basis to basis. It may be possible to ignore rotation invariance with these interpolations but there isn't a current implementation, primarily because the within-plane spin that is generated is distracting from a visualization perspective. Rotation invariance is checked for the proposed PPIs by rotating the structured projection, within the plane. <!--

@@

- **speed**: Being fast to compute allows the index to be used in real-time in a guided tour, where the optimization can be watched. When the computations are shifted off-line, to watch in replay, computation times matter less. This is checked  by comparing times for benchmark scenarios with varying sample size and dimension.

@@

Three families of data simulations are used for examining the behavior of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the PPIs on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $p$ is used for the number of points, $n$ is the number of dimensions, and $d=2$ is the projection dimension.

@@

The structured projection is in variables 5 and 6 ($x_5, x_6$). Two samples sizes are used: $n=(100, 1000)$. All variables are standardized to have mean 0 and standard deviation 1. Figure \ref{fig:data} shows samples from each of the families, of the nuisance and structured pairs of variables. Table \ref{tab:indexTable} compares the PPIs for structured projections against those for nuisance variables, based on 100 simulated data sets of each type, using sample size 1000. The lower and upper show the 5th and 95th percentile of values. All indexes, except convex show distinctly higher values for the structured projections. The convex index shows the inverse scale to other indices, thus (1-convex) will be used in the assessment of performance of PPIs. The scale for the holes index is smaller than the others ranging from about 0.7 through 1, so it is re-scaled in the performance assessment so that all indices can be plotted on a common scale of 0-1 (details in Appendix). Similarly, the TIC index is re-scaled.

@@

<!-- - Compare the index values for structured vs nuisance variables, by simulating many examples from the model families, and recording the 90% confidence intervals. This checks if the PPI can detect the structure, that it has distinct differences between the sets of values. 
-->
1. Compute the PPI values on the tour path along an interpolation between pairs of nuisance variables, $x_1 - x_2$ view either to the $x_3 - x_4$. The result is ideally a smooth change in low values. This checks the smoothness property. 
2. Change to a tour path between a pair of nuisance variables $x_1 - x_2$ and the structured pair of variables $x_5 - x_6$, and compute the PPI along this. This examines the squintability, and smoothness. If the function is smooth and slowly increases towards the structured projection, then the structure is visible from a distance. 
3. Use the guided tour to examine the ease of optimization. This depends on having a relatively smooth function, with structure visible from a distance. One index is optimized to show how effectively the maximum is attained, and the values for other PPIs is examined along the same path, to examine the similarity between PPIs. 
4. Rotation invariance is checked by computing PPIs on rotations of the structured projection. 
5. Some PPIs have a choice of parameters, and the choice can have an effect on function smoothness, and sensitivity to structure. The parameters for these PPIs is varied and examined along a tour path between noise and structured projections, to examine the effects of parameter choices and how fiddly the index is to tune. 
6. Computational speed for the selected indices is examined on a range of sample sizes.

@@

Figure \ref{fig:plannedtour} shows the PPIs for a tour sequence between a nuisance and structured projection. Sample size is indicated by line type: dashed being $n=100$ and solid is $n=1000$. The beginning of the sequence is the nuisance projection and the end is the structured projection. The index values for most PPIs increases substantially nearing the structured projection, indicating that they "see" the structure. Some indexes see all three structures: scagnostics, MIC and TIC,  which means that they are flexible indexes capable of detecting a range of structure. [@Grimm2016]'s indexes, cor2d and splines2d, are excellent for detecting the sine, and they can see it from far away, indicated by the long slow increase in index value. The holes index easily detects the pipe, and can see it from a distance. The scagnostic index, stringy, can see the structure but is myopic, only when it is very close. Interestingly the scagnostic, skinny, sees the spiral from a distance.

@@

The guided tour combines optimization with interpolation between pairs of planes. Target planes of the path are chosen to maximize the PPI. There are three derivative-free optimization methods available in the guided tour: \verb# search_better_random# (1), \verb# search_better# (2), and \verb# search_geodesic# (3). Method 1 casts a wide net randomly generating projection planes, computing the PPIs and keeping the best projection, and method 2 conducts a localized maximum search. Method 3 is quite different: a local search is conducted to determine a promising direction, and then this direction is followed until the maximum in that direction is found. For all methods the optimization is iterative, the best projections form target planes in the tour, the tour path is the interpolation to this target, and then a new search for a better projection is made, followed by the interpolation. For each projection during the interpolation steps, the PPI is recorded.

The stopping rule is that no better projections are found after a fixed number of tries, given a fixed tolerance value measuring difference. For method 1 and 2 two additional parameters control the optimization: the search window $\alpha$, giving the maximum distance from the current plane in which projections are sampled, and the cooling factor, giving the incremental decrease in search window size. Method 3 in principle also has two free parameters, which are however fixed in the current implementation. The first is the small step size used when evaluating the most promising direction, it is fixed to 0.01, and the second parameter being the window over which the line search is performed, fixed to $\pm \pi/4$ away from the current plane.

For distributions and indexes with smooth behavior and good squintability, method 3 is the most effective method for optimization. If these two criteria are not met the method may still be useful, but only given an informed starting projection. In such cases we can follow a method similar to that proposed by Friedman (XXX REF): we break the optimization in two distinct steps. A first step ("scouting") uses method 2 with large search window and no cooling as a way of stepping over fluctuations and local maxima and yielding an approximation of the global maximum. Note that this likely requires large number of tries, especially as dimension increases, since most randomly picked planes will not be interesting. The second step is uses method 3 starting from the approximate maximum, which will take small steps to refine the result to be closer to the global maximum.

@@

Despite the simple structure, the pipe is relatively difficult for the PPIs to find. For the TIC index, there is a fairly small squint angle. For the holes index, there are several local maxima, that divert the optimizer. There is a hint of this from Figure \ref{fig:plannedtour} because the initial projection (left side of trace) of purely noise variables has a higher index value than the linear combinations of noise and structured variables along the path. The uniform distribution was used to generate the noise variables, which has a higher PPI value than a normal distribution, yielding the higher initial value. In addition, a local maximum is observed whenever the pair of variables is one structured variable and one noise variable, because there is a lighter density in the center of the projection.
